---
title: "MATH3030 Coursework"
subtitle: "Multivariate Statistics --- Spring 2026"
author: "Aayush Parekh (20527121)"
date: ""
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: pdflatex
    fig_caption: yes
fontsize: 11pt
geometry: margin=2.5cm
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{microtype}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  dpi        = 200,
  cache      = TRUE,
  fig.pos    = "H",
  out.width  = "100%"
)
user_lib <- file.path(path.expand("~"), "R", "library")
.libPaths(c(user_lib, .libPaths()))
suppressPackageStartupMessages({
  library(ggplot2); library(ggrepel); library(MASS)
  library(mclust); library(patchwork); library(scales); library(knitr)
})
set.seed(42)

palette_continents <- c(
  "Africa"="#66c2a5","Americas"="#fc8d62","Asia"="#8da0cb",
  "Europe"="#e78ac3","Oceania"="#a6d854"
)
theme_paper <- theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 10))

UN <- read.csv("UN.csv")
years       <- seq(1952, 2007, 5)
year_labels <- as.character(years)
continent   <- UN[, 1]
country     <- UN[, 2]
gdp     <- UN[,  3:14]; colnames(gdp)     <- year_labels; rownames(gdp)     <- country
lifeExp <- UN[, 15:26]; colnames(lifeExp) <- year_labels; rownames(lifeExp) <- country
popn    <- UN[, 27:38]; colnames(popn)    <- year_labels; rownames(popn)    <- country
```

\newpage

# Question 1

## Exploratory Data Analysis

The UN dataset contains GDP per capita, life expectancy, and population
for 141 countries at 12 evenly spaced time points from 1952 to 2007.
No values are missing. Sample sizes per continent are uneven (52
African, 32 Asian, 30 European, 25 American, and 2 Oceanian), so any
continent-level summary for Oceania is effectively a two-country
average of Australia and New Zealand. We summarise the 70-year
dynamics with continent-mean trajectories for each variable and a
table of 1952 and 2007 medians.

```{r eda-trends, fig.cap="Continent-mean trajectories, 1952–2007. GDP and population are on log scales; life expectancy is linear.", fig.width=10, fig.height=3.2}
eda_long <- function(data, varname) {
  mat <- as.matrix(data)
  do.call(rbind, lapply(sort(unique(continent)), function(cont) {
    m_year <- colMeans(mat[continent == cont, , drop = FALSE])
    data.frame(year = years, continent = cont, value = as.numeric(m_year),
               variable = varname, stringsAsFactors = FALSE)
  }))
}
eda_trends <- rbind(
  eda_long(gdp,     "GDP per capita (USD)"),
  eda_long(lifeExp, "Life expectancy (years)"),
  eda_long(popn,    "Population")
)
panels <- lapply(c("GDP per capita (USD)","Life expectancy (years)","Population"),
  function(v) {
    d <- subset(eda_trends, variable == v)
    p <- ggplot(d, aes(year, value, colour = continent)) +
      geom_line(linewidth = 0.9) + geom_point(size = 1.3) +
      scale_colour_manual(values = palette_continents) +
      labs(x = "Year", y = NULL, title = v) + theme_paper
    if (v != "Life expectancy (years)") p <- p + scale_y_log10(labels = label_comma())
    else p <- p + scale_y_continuous(labels = label_comma())
    p
  })
wrap_plots(panels, ncol = 3, guides = "collect") &
  theme(legend.position = "right") &
  labs(colour = "Continent")
```

Asia shows the steepest gains on both axes. Mean life expectancy rose
from 44 to 72 years and median GDP per capita almost quadrupled,
driven by a handful of export-led economies. Singapore's GDP per
capita rose from \$2,315 to \$47,143 (20×), South Korea's by 23×,
and Taiwan's by 24×. China dominates the population panel, rising
from 556 million in 1952 to 1.32 billion in 2007. These multipliers
are well above the Asian median GDP gain of 3.8×, so the continent
mean hides a wide spread rather than uniform growth.

Africa's trajectory is much shallower. Median GDP rose only 1.5× over
55 years, and mean life expectancy, after rising from 39 to 53 years,
plateaued from the 1990s onward. The plateau coincides with a broad
health and conflict downturn in southern and central Africa. Zimbabwe's life expectancy fell from 48.5 to 43.5 years
(the largest decline in the dataset), Swaziland, Zambia, and Lesotho
show similar stagnation, and the Democratic Republic of the Congo is
the only country whose GDP per capita ended below its 1952 level, at
36% of its original value. Africa's population grew faster than any
other continent's, which compounds the GDP-growth shortfall on a
per-capita basis.

Europe and Oceania remain firmly at the top throughout. Europe's GDP
growth (\$5,143 to \$28,054, 5.5×) is comparable to Asia's in
multiplier terms, but its life-expectancy gain of +13 years is about
half of Asia's +28. This is structural rather than a lack of
development. A population already close to its demographic ceiling
has less room to gain years than one that is catching up. Both
continents also evolve more smoothly than Asia or Africa, with no
sharp deviations between adjacent five-year intervals.

```{r eda-bubble, fig.cap="GDP per capita vs life expectancy in 2007. Bubble size is proportional to population.", fig.width=7, fig.height=4.8}
bubble_df <- data.frame(country = country, continent = continent,
                        gdp = gdp[,"2007"], lifeExp = lifeExp[,"2007"],
                        popn = popn[,"2007"])
largest <- do.call(rbind, by(bubble_df, bubble_df$continent,
  function(d) d[which.max(d$popn), ]))
ggplot(bubble_df, aes(gdp, lifeExp, size = popn, colour = continent)) +
  geom_point(alpha = 0.6) +
  geom_text_repel(data = largest, aes(label = country),
                  size = 3, fontface = "bold", show.legend = FALSE) +
  scale_x_log10(labels = label_comma()) +
  scale_size_area(max_size = 15, guide = "none") +
  scale_colour_manual(values = palette_continents) +
  labs(x = "GDP per capita (log scale, 2007)",
       y = "Life expectancy (years, 2007)", colour = "Continent") +
  theme_paper
```

The 2007 snapshot compresses the trajectories into a single
cross-section and makes the joint structure visible. GDP and life
expectancy are strongly but non-linearly related. The Pearson
correlation on log(GDP) is 0.81, appreciably higher than on raw GDP
(0.68), reflecting the diminishing marginal returns on life expectancy
at high income. Continents separate cleanly, with Africa in the
lower-left, Europe in the upper-right, and the Americas grouped in the
middle. Asia is the outlier continent, spanning almost
the full income range from India and Bangladesh at the low end to
Japan and Singapore at the top, with China positioned between the two
modes.

```{r eda-kde, fig.cap="Kernel density estimates of each variable in 2007.", fig.width=11, fig.height=3.2}
kde_specs <- list(
  list(name="log10(GDP per capita, 2007)", x=log10(gdp[,"2007"])),
  list(name="Life expectancy (2007)",      x=lifeExp[,"2007"]),
  list(name="log10(Population, 2007)",     x=log10(popn[,"2007"])))
kde_panels <- lapply(kde_specs, function(s) {
  d <- data.frame(x = s$x, continent = continent)
  ggplot(d, aes(x, fill = continent, colour = continent)) +
    geom_density(alpha = 0.2, linewidth = 0.7) +
    scale_fill_manual(values = palette_continents, guide = "none") +
    scale_colour_manual(values = palette_continents) +
    labs(x = NULL, y = "Density", title = s$name, colour = "Continent") +
    theme_paper + theme(axis.text.y = element_blank(),
                        axis.ticks.y = element_blank())
})
wrap_plots(kde_panels, ncol = 3, guides = "collect") &
  theme(legend.position = "right")
```

The KDEs show the within-continent distributional shape. Asia's GDP
distribution is visibly bimodal (Gulf states and East Asian tigers on
one side, South and Central Asia on the other), with an IQR ratio of
8.9, the largest of any continent. Africa's life-expectancy
distribution is wide and left-skewed, with a lower tail formed by
Swaziland, Zambia, Zimbabwe, and Lesotho, all below 45 years.
Population densities peak near 10 million in every continent, with
Asia carrying the long right tail that contains China and India.
Continents that appeared internally homogeneous on the trajectory
plot thus conceal real heterogeneity under the mean.

```{r eda-table}
summary_tbl <- data.frame(
  N          = as.vector(table(continent)),
  `GDP 1952` = round(tapply(gdp[,"1952"],     continent, median), 0),
  `GDP 2007` = round(tapply(gdp[,"2007"],     continent, median), 0),
  `LE 1952`  = round(tapply(lifeExp[,"1952"], continent, median), 1),
  `LE 2007`  = round(tapply(lifeExp[,"2007"], continent, median), 1),
  `Pop 2007 (M)` = round(tapply(popn[,"2007"], continent, sum) / 1e6, 0),
  check.names = FALSE
)
rownames(summary_tbl) <- sort(unique(continent))
kable(summary_tbl,
      caption = "Median GDP per capita (USD) and life expectancy (years) by continent; population (millions).",
      booktabs = TRUE)
```

A small number of countries deviate strongly from their continent's
trend and reappear as extreme points in the PCA, CCA, and MDS
projections below. Equatorial Guinea's GDP per capita rose
32× on the back of offshore oil discoveries, Zimbabwe's life
expectancy fell by five years, Singapore, South Korea, and Taiwan
achieved sustained 20–24× GDP multiples, and the Democratic Republic
of the Congo and Haiti are the two economies that regressed most
sharply, ending the period at 36% and 65% of their 1952 GDP per
capita respectively.

## Principal Component Analysis

We perform PCA independently on the $141 \times 12$ GDP and
life-expectancy blocks. GDP values span more than two orders of
magnitude across countries, so we standardise each year column (which
is equivalent to decomposing the correlation matrix) to prevent
larger recent-year variances from dominating. Life expectancy is on a
comparable scale across years, but we also standardise it. The
covariance and correlation decompositions produce near-identical
results (PC1 explains 92.0% and 92.3% of the variance respectively),
and standardising keeps the treatment of the two variable blocks
consistent.

```{r pca-fit}
pca_gdp  <- prcomp(gdp,     scale. = TRUE)
pca_life <- prcomp(lifeExp, scale. = TRUE)
var_ratio <- function(p) (p$sdev^2) / sum(p$sdev^2)
```

### Scree analysis and variance explained

```{r pca-scree, fig.cap="Scree plots for the GDP and life-expectancy PCAs.", fig.width=11, fig.height=4}
scree_df <- function(p, label) data.frame(pc = seq_along(var_ratio(p)),
  var = var_ratio(p), cum = cumsum(var_ratio(p)), which = label)
scree <- rbind(scree_df(pca_gdp, "GDP per capita"),
               scree_df(pca_life, "Life expectancy"))
ggplot(scree, aes(pc)) +
  geom_col(aes(y = var), fill = "steelblue", alpha = 0.6) +
  geom_line(aes(y = cum), colour = "darkred") +
  geom_point(aes(y = cum), colour = "darkred", size = 1.3) +
  geom_hline(yintercept = 0.9, linetype = "dashed",
             colour = "grey40", linewidth = 0.3) +
  facet_wrap(~which) +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Principal component", y = "Proportion of variance") +
  theme_paper
```

Both decompositions concentrate almost all variance in a single
component. PC1 explains 91.8% of the GDP variance and 92.3% of the
life-expectancy variance, while PC2 adds a further 5.3% and 5.6%
respectively. Two components are sufficient, together covering 97.1%
of the GDP variance and 97.9% of the life-expectancy variance. This
follows directly from the year-to-year redundancy within each block,
with adjacent years correlated at approximately 0.99, so 12 columns
compress cleanly into two.

### Interpretation of PC1 and PC2

The loadings have the same structure on both decompositions. PC1
assigns nearly uniform positive weights across all 12 years (0.28 to
0.30 on GDP, 0.27 to 0.30 on life expectancy), so a country's PC1
score is approximately proportional to its average level over the
full 55-year period. For PC2, the loadings begin strongly negative in
1952 ($-0.35$) and rise monotonically to strongly positive in 2007
($+0.42$). A positive PC2 score therefore indicates a country whose
values were low early and high late (a catching-up or fast-growth
trajectory), and a negative PC2 score indicates the opposite
(stagnation or decline from a relatively high base).

### PC score scatter plots

```{r pca-align, include=FALSE}
sw <- which(country == "Switzerland")
if (length(sw) && pca_gdp$x[sw, 1] < 0)  pca_gdp$x[, 1]  <- -pca_gdp$x[, 1]
if (length(sw) && pca_life$x[sw, 1] < 0) pca_life$x[, 1] <- -pca_life$x[, 1]
sg <- which(country == "Singapore")
if (length(sg) && pca_gdp$x[sg, 2] < 0) pca_gdp$x[, 2] <- -pca_gdp$x[, 2]
om <- which(country == "Oman")
if (length(om) && pca_life$x[om, 2] < 0) pca_life$x[, 2] <- -pca_life$x[, 2]
```

```{r pca-scores, fig.cap="PC1 vs PC2 scatter plots for GDP (left) and life expectancy (right), coloured by continent.", fig.width=13, fig.height=5}
scatter_df <- function(p, label) {
  d <- data.frame(PC1 = p$x[,1], PC2 = p$x[,2], country = country,
                  continent = continent, which = label)
  d$dist <- sqrt(d$PC1^2 + d$PC2^2); d
}
make_scatter <- function(d, title) {
  top <- d[order(-d$dist), ][1:8, ]
  ggplot(d, aes(PC1, PC2, colour = continent)) +
    geom_point(size = 1.8, alpha = 0.75) +
    geom_text_repel(data = top, aes(label = country),
                    size = 2.3, show.legend = FALSE, max.overlaps = Inf) +
    scale_colour_manual(values = palette_continents) +
    labs(title = title, colour = "Continent") + theme_paper
}
(make_scatter(scatter_df(pca_gdp,  "GDP"), "GDP — PC1 vs PC2") |
 make_scatter(scatter_df(pca_life, "Life Expectancy"), "Life Expectancy — PC1 vs PC2")) +
  plot_layout(guides = "collect") & theme(legend.position = "right")
```

The GDP scatter is structured by level on PC1. Switzerland, the US,
Norway, Canada, and Denmark sit at the top end, while Myanmar,
Burundi, Ethiopia, Eritrea, and Malawi sit at the bottom. PC2 then
separates countries by growth trajectory. The top Asian countries
form the obvious cluster at the top of PC2 (Singapore, Hong Kong,
Taiwan, South Korea, Japan), identifying the countries whose relative
position improved most over the period. Strongly negative PC2 scores
belong to Venezuela, Iraq, and Libya, relatively high-income
economies in 1952 that did not sustain their early position
advantage.

The life-expectancy scatter has the same geometry. PC1 separates
Iceland, Sweden, Norway, the Netherlands, and Switzerland at the top
and Sierra Leone, Afghanistan, Angola, Guinea-Bissau, and Mozambique
at the bottom. PC2 captures the rate of improvement, with Oman,
Vietnam, Saudi Arabia, Indonesia, and Libya gaining life expectancy
fastest, while Zimbabwe, Zambia, Rwanda, Botswana, and the Democratic
Republic of the Congo have the strongly negative PC2 scores
consistent with the health and conflict downturn identified in the
EDA. Continent clusters are distinct, with Europe and Oceania
at the top of PC1, Africa at the bottom, the Americas in the middle,
and Asia stretched along PC2 consistent with its bimodal GDP
distribution.

### GDP PC1 versus life-expectancy PC1

```{r pca-cross, fig.cap="First PC scores, GDP against life expectancy.", fig.width=7, fig.height=5.5}
cross <- data.frame(gdp_pc1 = pca_gdp$x[,1], life_pc1 = pca_life$x[,1],
                    country = country, continent = continent)
cross$dist <- sqrt(cross$gdp_pc1^2 + cross$life_pc1^2)
ggplot(cross, aes(gdp_pc1, life_pc1, colour = continent)) +
  geom_point(size = 2, alpha = 0.75) +
  geom_text_repel(data = cross[order(-cross$dist), ][1:10, ],
                  aes(label = country), size = 2.4, show.legend = FALSE) +
  scale_colour_manual(values = palette_continents) +
  labs(x = "GDP PC1 score", y = "Life Expectancy PC1 score", colour = "Continent") +
  theme_paper
```

The two PC1 scores plotted against each other show how economic
level and health level covary over the full period. The Pearson
correlation is `r round(cor(pca_gdp$x[,1], pca_life$x[,1]), 2)`,
consistent with the 0.81 log(GDP)–life expectancy correlation seen
in the 2007 snapshot and confirming the long-run coupling between
wealth and health. Most countries lie close to the diagonal. The systematic
deviations are countries whose health outcomes are either stronger or
weaker than their income level would predict, reflecting external
factors such as public health investment, diet, and health system
access that the economic dimension alone does not capture.

### Limitations of linear PCA

The decomposition is restricted to linear combinations of the year
columns and the GDP–life expectancy relationship is non-linear
(Pearson correlation 0.68 on raw GDP against 0.81 on log(GDP)).
Kernel PCA, which applies the standard decomposition in a feature
space induced by a kernel, would capture this curvature directly and
would be expected to produce tighter continent separation than the
linear projection used here.^[Hastie, Tibshirani, and Friedman,
*The Elements of Statistical Learning* (2nd ed., 2009), Chapter 14.]

## Canonical Correlation Analysis

We apply CCA with $\log(\text{GDP})$ as the first variable set and
life expectancy as the second, giving $p = q = 12$. The log transform
is motivated by the same skewness and non-linearity noted in the EDA
and the PCA limitations subsection. Raw GDP is strongly right-skewed
(skewness 1.88) and the 2007 GDP–life expectancy relationship curves
noticeably on the raw scale but is close to linear on the log scale.

```{r cca-fit}
log_gdp  <- log(as.matrix(gdp))
life_mat <- as.matrix(lifeExp)
cc  <- cancor(log_gdp, life_mat)
cc_raw <- cancor(as.matrix(gdp), life_mat)
U <- scale(log_gdp,  scale = FALSE) %*% cc$xcoef
V <- scale(life_mat, scale = FALSE) %*% cc$ycoef
```

```{r cca-scatter, fig.cap="First pair of canonical variables, coloured by continent.", fig.width=7, fig.height=5.5}
cca_df <- data.frame(U1 = U[,1], V1 = V[,1],
                     country = country, continent = continent)
cca_df$dist <- sqrt(cca_df$U1^2 + cca_df$V1^2)
ggplot(cca_df, aes(U1, V1, colour = continent)) +
  geom_point(size = 2, alpha = 0.75) +
  geom_text_repel(data = cca_df[order(-cca_df$dist), ][1:8, ],
                  aes(label = country), size = 2.4, show.legend = FALSE) +
  scale_colour_manual(values = palette_continents) +
  labs(x = "First CC variable (log GDP side)",
       y = "First CC variable (life expectancy side)",
       title = sprintf("CCA: First canonical pair  (r = %.3f)", cc$cor[1]),
       colour = "Continent") +
  theme_paper
```

### Canonical correlations

The first canonical correlation is `r round(cc$cor[1], 2)`, and
subsequent correlations decay to `r round(cc$cor[2], 2)`,
`r round(cc$cor[3], 2)`, and `r round(cc$cor[4], 2)`, with the
remainder below 0.40. The first pair therefore carries most of the
shared linear structure between the two blocks. Looking at CC2
directly, the GDP-side loadings start near zero in 1952 (0.003), rise
to 0.11 around 1977, and fall back to $-0.11$ by 2007, while the
life-expectancy loadings are uniformly small and negative across all
12 years ($-0.05$ to $-0.19$). The GDP pattern alone describes a
faint mid-period growth contrast, but it is not matched by any
equivalent temporal structure on the life-expectancy side, so the two
variates of CC2 are not aligned on the same temporal feature and the
pair has no joint interpretation. Higher-order pairs decay further
and are similarly uninterpretable.

### Interpretation of the first canonical pair

The first canonical axis ranks countries by a joint development level
that log(GDP) and life expectancy identify together. The lower
extreme is dominated by sub-Saharan African and conflict-affected
Asian states, with Mozambique, Myanmar, and Sierra Leone at the
bottom, and the upper extreme by Western European countries and the
United States, with Norway and Switzerland at the top. This ordering
is near-identical to the one PCA produced on either variable
separately, because the 12 year columns within each block are
correlated at approximately 0.99 and there is little room for CCA's
maximum-correlation objective to choose a direction meaningfully
different from PCA's maximum-variance one. The continent structure on
the canonical scatter reproduces the PCA pattern, with Africa in the
lower-left and Europe and Oceania in the upper-right.

### Effect of the log transform

Repeating the analysis on raw GDP gives a first canonical correlation
of `r round(cc_raw$cor[1], 2)`, materially lower than the
`r round(cc$cor[1], 2)` achieved with log(GDP). CCA only finds linear
combinations of the variables it is given, so any curvature in the
raw GDP–life expectancy relationship is left unaccounted for. The
log transform straightens the relationship and raises the canonical
correlation. The gain is the PCA-section's linearity limitation
measured on a different axis.

## Multidimensional Scaling

Classical (metric) MDS is applied to the combined 36-dimensional
dataset formed by concatenating log(GDP), life expectancy, and
log(population), using Euclidean distance as specified in the
coursework. The log transforms on GDP and population bring them onto
a roughly comparable scale with life expectancy.

```{r mds-fit}
UN_transformed <- cbind(log(as.matrix(gdp)), as.matrix(lifeExp), log(as.matrix(popn)))
d_full    <- dist(UN_transformed)
mds       <- cmdscale(d_full, k = 2, eig = TRUE)
d_reduced <- dist(mds$points)
dist_cor  <- cor(as.vector(d_full), as.vector(d_reduced))
```

The Pearson correlation between pairwise distances in the original
36-D space and the 2D MDS projection is
`r round(dist_cor, 3)`, so the two retained dimensions preserve the
distance geometry of the full data almost exactly.

```{r mds-plot, fig.cap="Two-dimensional MDS representation, coloured by continent.", fig.width=8, fig.height=6}
mds_df <- data.frame(D1 = mds$points[,1], D2 = mds$points[,2],
                     country = country, continent = continent)
mds_df$dist <- sqrt(mds_df$D1^2 + mds_df$D2^2)
ggplot(mds_df, aes(D1, D2, colour = continent)) +
  geom_point(size = 2, alpha = 0.75) +
  geom_text_repel(data = mds_df[order(-mds_df$dist), ][1:10, ],
                  aes(label = country), size = 2.4, show.legend = FALSE) +
  scale_colour_manual(values = palette_continents) +
  labs(x = "MDS dimension 1", y = "MDS dimension 2", colour = "Continent") +
  theme_paper
```

### Development axis

Dimension 1 is a clear development axis. The continent centroids line
up from Africa at $-37$ through the Americas at $+18$ and Europe at
$+44$ to Oceania at $+52$, in the same order they occupy on every
previous projection. The lower extreme is dominated by sub-Saharan
African and conflict-affected Asian states such as Sierra Leone,
Afghanistan, Angola, Guinea-Bissau, and Rwanda, while the upper
extreme is occupied by Scandinavian and Alpine European countries
including Sweden, Iceland, Norway, the Netherlands, and Switzerland.
This country ordering matches the one produced by PCA and CCA, since
all three methods are projecting the same dominant development signal
that accounts for the bulk of the variation in the data.

### Growth direction

Dimension 2 separates countries by the direction of change in life
expectancy over the period rather than by level. The most negative
scores belong to countries with the largest life-expectancy gains
over the 55 years, a cluster containing Oman, Vietnam, Indonesia, and
Yemen, while the most positive scores belong to countries whose life
expectancy fell or stagnated over this period, including Zimbabwe,
Botswana, Zambia, and Swaziland. Asia's centroid is pulled
below the horizontal axis by its concentration of fast-gaining
countries, while Africa's centroid sits near the horizontal because
the continent contains both rapidly improving and declining cases.
The contrast is the same one that appeared as PC2 in the
life-expectancy PCA, now reproduced with all three variables analysed
simultaneously.

### Relationship to PCA

The agreement between the MDS and PCA axes is not a coincidence.
Classical MDS with Euclidean distance is, up to sign, the same
decomposition as PCA on the same centered data. Writing the SVD of
the centered data matrix as
$$X = U D V^T,$$
PCA returns the scores
$$T_{\text{PCA}} = X V = U D.$$
Classical MDS forms the Gram matrix $B = X X^T$ and takes the
eigendecomposition
$$B = U D^2 U^T,$$
from which the MDS coordinates are the leading columns of $U D$, the
same matrix as $T_{\text{PCA}}$. The MDS scatter does not literally
reproduce either PCA scatter because the input differs. Here MDS is
applied to the 36-column concatenation of $\log(\text{GDP})$, life
expectancy, and $\log(\text{popn})$, whereas the PCA sections ran the
decomposition on the GDP and life-expectancy blocks separately.

### Comparison with the earlier analyses

The MDS projection recovers the same two axes as the individual PCAs,
with dimension 1 tracking development level and dimension 2 tracking
growth direction. The MDS version adds a third variable by operating
on all three simultaneously, so a single 2D summary captures joint
structure that required two separate PCAs above. The 0.997
correlation between pairwise distances in the original 36-D space and
the 2D MDS projection makes the reduced representation a near-lossless
summary, so the development and growth-direction axes are the two
dominant sources of variation in the full UN data.

\newpage

# Question 2

## Linear Discriminant Analysis

We use LDA to predict continent from all 36 features (GDP, life
expectancy, and population across 12 years). The data are split 70/30
into training and test sets.

```{r lda-fit}
X_all <- as.matrix(cbind(gdp, lifeExp, popn))
y_all <- factor(continent)
set.seed(42)
n <- nrow(X_all)
idx <- sample(n, round(0.7 * n))
fit_lda <- MASS::lda(X_all[idx, ], grouping = y_all[idx])
pred    <- predict(fit_lda, X_all[-idx, ])$class
acc_single <- mean(pred == y_all[-idx])
cm <- table(True = y_all[-idx], Pred = pred)

accs <- numeric(50)
for (s in 1:50) {
  set.seed(s)
  i_tr <- sample(n, round(0.7 * n))
  m    <- MASS::lda(X_all[i_tr, ], grouping = y_all[i_tr])
  p    <- predict(m, X_all[-i_tr, ])$class
  accs[s] <- mean(p == y_all[-i_tr])
}
```

A single fixed split with seed 42 gives a test accuracy of
`r sprintf("%.3f", acc_single)` on `r nrow(X_all) - length(idx)`
held-out countries, and averaging over 50 random splits gives a more
stable estimate of `r sprintf("%.3f ± %.3f", mean(accs), sd(accs))`.
Both figures sit comfortably above the 0.20 five-class chance rate
and above the 0.37 majority-class baseline set by Africa's sample
proportion, so the classifier is extracting genuine structure from
the features.

```{r lda-confusion}
kable(as.matrix(cm),
      caption = "Rows are true continents, columns are predicted.",
      booktabs = TRUE)
```

Per-class accuracy varies substantially. Africa is the best-classified
continent, followed by Europe, Asia, and the Americas. The Oceanian
country in the test set is misclassified as European. Asia's errors
are spread across Africa, the Americas, and Europe in roughly equal
proportions.

The per-class pattern is consistent with LDA's modelling assumptions.
The classifier assigns country $i$ to the class $c$ that maximises
$$\delta_c(x) = x^\top \Sigma^{-1} \mu_c - \tfrac{1}{2} \mu_c^\top \Sigma^{-1} \mu_c + \log \pi_c,$$
where $\mu_c$ is the class mean, $\Sigma$ is the pooled within-class
covariance, and $\pi_c$ is the class prior. The decision rule is
therefore linear in $x$, and the two modelling assumptions behind it
are that each class is multivariate normal and that all classes share
the same covariance $\Sigma$. Both assumptions break on this dataset.
The EDA showed Asia's GDP distribution is strongly bimodal, so Asia's
single mean $\mu_{\text{Asia}}$ lands between its two modes and the
density $N(\mu_{\text{Asia}}, \Sigma)$ concentrates mass in a region
where almost no Asian country actually sits, while overlapping with
the Gaussian densities for Africa, the Americas, and Europe. The
shared covariance is similarly misspecified because Oceania and
Europe have far tighter within-class spread than Africa or Asia, so
$\Sigma$ is an average that fits neither. Africa is classified more
accurately than Asia despite being internally heterogeneous because
its spread runs along a single unimodal axis, which the Gaussian
model can represent even if the variance is wrong.

## Clustering

Three methods from the module are applied to the standardised GDP and
life-expectancy blocks (24 features). k-means and the model-based
clustering are fitted directly, and hierarchical clustering is
applied via its linkage-distance dendrogram, using Ward's linkage as
in the computer lab sessions. Ward merges the pair of clusters
producing the smallest increase in within-cluster variance at each
step. $k$ is chosen from the dendrogram and the MVN-mixture BIC
curve, with the same $k$ then used for k-means so that all three
methods can be compared on equal footing.

```{r clust-fit}
X_cluster <- scale(cbind(as.matrix(gdp), as.matrix(lifeExp)))
h         <- hclust(dist(X_cluster), method = "ward.D2")
mc_models <- Mclust(X_cluster, G = 1:8, verbose = FALSE)
bic_per_G <- apply(mc_models$BIC, 1, function(v) max(v, na.rm = TRUE))
```

```{r clust-selection, fig.cap="Hierarchical clustering dendrogram (left) and model-based BIC as a function of number of components (right).", fig.width=11, fig.height=4}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(h, labels = FALSE, main = "Hierarchical clustering (Ward)",
     xlab = "Countries", sub = "", ylab = "Linkage distance")
plot(as.numeric(names(bic_per_G)), bic_per_G, type = "b",
     xlab = "Number of components G", ylab = "BIC (best model)",
     main = "Model-based (mclust) BIC", col = "steelblue", pch = 19)
```

### Choice of k

The dendrogram's three largest merge distances are
`r sprintf("%.1f, %.1f, and %.1f", sort(h$height, decreasing=TRUE)[1], sort(h$height, decreasing=TRUE)[2], sort(h$height, decreasing=TRUE)[3])`,
with every subsequent merge below 16. The substantial gap between
merges 3 and 4 identifies $k = 3$ as the cleanest cut point from the
hierarchical perspective. The MVN-mixture BIC curve has a clear
maximum at $G = 2$ (BIC `r round(bic_per_G[2], 0)`), with $G = 3$
second-best at BIC `r round(bic_per_G[3], 0)` and $G \geq 4$ falling
sharply. The two selection criteria therefore disagree, with the
dendrogram preferring $k = 3$ and the mixture model preferring
$k = 2$. We proceed with $k = 3$ because the BIC gap between $G = 2$
and $G = 3$ is moderate relative to the drop-off from $G = 3$ to
$G = 4$, and the three-cluster partition is the more informative one
against the development-axis structure recovered by PCA and MDS.

### Agreement between methods

```{r clust-compare}
best_k <- 3
km_res   <- kmeans(X_cluster, centers = best_k, nstart = 10)
hier_res <- cutree(h, k = best_k)
mc_res   <- Mclust(X_cluster, G = best_k, verbose = FALSE)$classification
ari_km_h  <- adjustedRandIndex(km_res$cluster, hier_res)
ari_km_mc <- adjustedRandIndex(km_res$cluster, mc_res)
ari_h_mc  <- adjustedRandIndex(hier_res,       mc_res)
```

```{r clust-comparison, fig.cap="Cluster assignments at k = 3 for k-means, hierarchical clustering, and the MVN mixture.", fig.width=13, fig.height=4}
pca2 <- prcomp(X_cluster)$x[, 1:2]
cdf <- data.frame(PC1 = pca2[,1], PC2 = pca2[,2],
                  kmeans = factor(km_res$cluster),
                  hier   = factor(hier_res),
                  mvn    = factor(mc_res),
                  continent = continent)
pl <- function(lab, ti) ggplot(cdf, aes(PC1, PC2, colour = .data[[lab]])) +
  geom_point(size = 1.8, alpha = 0.8) + labs(title = ti, colour = lab) +
  theme_paper
(pl("kmeans", "k-means (k=3)") | pl("hier", "Hierarchical (k=3)") |
 pl("mvn", "MVN mixture (G=3)")) &
  theme(legend.position = "none")
```

At $k = 3$ the Adjusted Rand Index is `r round(ari_km_h, 2)` between
k-means and hierarchical clustering, `r round(ari_km_mc, 2)` between
k-means and the MVN mixture, and `r round(ari_h_mc, 2)` between
hierarchical and the MVN mixture. The k-means and hierarchical
results are nearly identical, consistent with both being
variance-minimising methods on the same standardised features. The
MVN mixture disagrees more because mclust fits full-rank covariance
structures (varying shape, equal volume) rather than the spherical
within-cluster geometry that k-means and Ward's method assume, and
the covariance flexibility lets it draw cluster boundaries that
follow the anisotropic spread of the data rather than minimising
within-cluster variance.

### Cluster interpretation

The three clusters have a natural development interpretation. The
first cluster contains 18 of the 30 European countries, 6 Asian
countries, and both Oceanian countries, placing it at the high-GDP
and high-life-expectancy end of the development axis already seen in
PCA and MDS. The second cluster contains 42 of the 52 African
countries together with 10 Asian and 2 American ones, and sits at the
low-development end. The third is the middle cluster, dominated by
the Americas (21 of 25) with a substantial Asian contingent (16 of
32) and the remaining 12 European countries. The ordering reproduces
the PCA and MDS development axis and shows that continent is a
reasonable but imperfect proxy for development level. Africa and
Europe fall almost entirely into the low and high clusters
respectively, while the Americas and Asia are split across the middle
and the extremes.

## Linear Regression

We predict 2007 life expectancy from the 11 GDP columns for 1952–2002,
excluding 2007 GDP to avoid contemporaneous leakage. The 11
predictors are correlated with one another at approximately 0.99 (the
year-to-year structure noted in the EDA), which is textbook
multicollinearity and motivates testing methods that handle it
directly rather than OLS alone. Six models are compared. Ordinary
least squares, ridge regression, and principal component regression
(PCR) are each applied to raw GDP and to $\log(\text{GDP})$.

Performance is measured by leave-one-out cross-validation (LOOCV).
Each country is held out in turn,
the model is refit on the remaining 140, and the held-out country is
predicted from that fit.^[Hastie, Tibshirani, and Friedman, *The
Elements of Statistical Learning* (2nd ed., 2009), Section 7.10.]
Writing $\hat y_i$ for the prediction made for country $i$ by the
model fit without it, the LOOCV score is
$$\text{RMSE}_{\text{LOOCV}} = \sqrt{\tfrac{1}{141} \sum_{i=1}^{141} (y_i - \hat y_i)^2}.$$
Each $\hat y_i$ is produced by a fit that never saw country $i$, so
the 141 residuals are genuine prediction errors rather than in-sample
fit.

Ridge and PCR have hyperparameters ($\alpha$ and $k$), which we tune
by nested cross-validation. The LOOCV above is the outer loop. Within
each outer fold, an inner 10-fold CV on the 140 training rows picks
$\alpha$ or $k$, and the selected value is used to refit on those 140
rows and predict the held-out country. The held-out country never
enters the inner loop, so it cannot influence the hyperparameter used
to score it. A single flat LOOCV that chose $\alpha$ or $k$ from all
141 residuals would feed each country's residual back into its own
hyperparameter selection, pulling the reported RMSE below the true
out-of-sample value. OLS has no hyperparameter and is unaffected.

### Ridge regression and principal component regression

Ridge regression minimises
$$\|y - X\beta\|^2 + \alpha \|\beta\|^2,$$
adding a penalty on the squared length of $\beta$ to the usual
least-squares loss. When predictors are near-collinear, OLS fits tend
to pair large positive coefficients with large negative ones that
mostly cancel on the fitted values. The ridge penalty rules these out
and returns a smaller, more stable set of coefficients in exchange
for a small amount of bias. The parameter $\alpha$ sets the trade-off.
At $\alpha = 0$ Ridge reduces to OLS, and as $\alpha \to \infty$ the
coefficients are driven to zero. We search $\alpha$ on a 50-point
log-spaced grid in $[10^{-2}, 10^3]$.

Principal component regression replaces the 11 near-collinear year
columns with their first $k$ principal components and fits OLS in
that reduced space. The PC directions beyond the leading few carry
almost no variance in the GDP block, so including them in OLS would
inflate the fit's sensitivity to them without supplying any real
signal. Cutting the number of components at $k$ eliminates those
directions outright.

### Ridge and PCR as shrinkage on principal directions

Both estimators can be written as filters on the principal directions
of the predictor matrix. Let the SVD of the centered $X$ be
$$X = U D V^T,$$
where $D$ holds the singular values in decreasing order. The OLS fit
projects $y$ onto the column space of $X$, weighting every PC
direction equally, while Ridge and PCR replace those unit weights
with component-wise filters,
$$\hat y_{\text{OLS}} = \sum_i u_i u_i^T y, \qquad
  \hat y_{\text{Ridge}} = \sum_i \frac{d_i^2}{d_i^2 + \alpha} u_i u_i^T y, \qquad
  \hat y_{\text{PCR}} = \sum_{i=1}^{k} u_i u_i^T y.$$
Ridge's factor is nearly one on high-variance directions and near
zero on low-variance ones, giving a smooth shrinkage. PCR is the hard
version, keeping the first $k$ directions untouched and dropping the
rest. In the GDP block the first three singular values dominate the
remaining eight by orders of magnitude, so both filters produce fits
built almost entirely from the first three PC directions. Ridge's
nested-CV RMSE and PCR's are two smoothings of the same low-rank
structure, explaining why they land so close.

```{r reg-setup}
y_reg <- lifeExp[, "2007"]
X_raw <- as.matrix(gdp[, 1:11]); colnames(X_raw) <- paste0("y", colnames(X_raw))
X_log <- log(X_raw)
alpha_grid <- 10^seq(-2, 3, length.out = 50)
k_grid     <- 1:11

ridge_fit <- function(X, y, alpha) {
  mx <- colMeans(X); sx <- apply(X, 2, sd); my <- mean(y)
  Xs <- sweep(sweep(X, 2, mx, "-"), 2, sx, "/"); ys <- y - my
  list(mx=mx, sx=sx, my=my,
       beta_s = solve(crossprod(Xs) + alpha * diag(ncol(Xs)), crossprod(Xs, ys)))
}
ridge_predict <- function(fit, Xnew) {
  Xs <- sweep(sweep(Xnew, 2, fit$mx, "-"), 2, fit$sx, "/")
  as.numeric(Xs %*% fit$beta_s) + fit$my
}
pcr_fit <- function(X, y, k) {
  mx <- colMeans(X); sx <- apply(X, 2, sd)
  Xs <- sweep(sweep(X, 2, mx, "-"), 2, sx, "/")
  p <- prcomp(Xs, scale. = FALSE); S <- p$x[, 1:k, drop = FALSE]
  list(mx=mx, sx=sx, rot=p$rotation[, 1:k, drop = FALSE],
       lm_fit=lm(y ~ ., data = data.frame(y=y, S)))
}
pcr_predict <- function(fit, Xnew) {
  Xs <- sweep(sweep(Xnew, 2, fit$mx, "-"), 2, fit$sx, "/")
  as.numeric(predict(fit$lm_fit, newdata = data.frame(Xs %*% fit$rot)))
}
loocv_ols <- function(X, y) {
  sapply(seq_along(y), function(i) {
    m <- lm(y ~ ., data = data.frame(y=y[-i], X[-i, ]))
    predict(m, newdata = data.frame(t(X[i, ])))
  })
}
rmse <- function(a, b) sqrt(mean((a - b)^2))
r2   <- function(a, b) 1 - sum((a - b)^2) / sum((a - mean(a))^2)
```

```{r reg-nested, cache=TRUE}
nested_ridge <- function(X, y, grid, n_inner = 10) {
  n <- length(y); preds <- numeric(n); alphas <- numeric(n)
  for (i in 1:n) {
    Xtr <- X[-i, ]; ytr <- y[-i]; set.seed(i)
    fold <- sample(rep(1:n_inner, length.out = nrow(Xtr)))
    cv <- sapply(grid, function(a) mean(sapply(1:n_inner, function(k) {
      fit <- ridge_fit(Xtr[fold != k, ], ytr[fold != k], a)
      rmse(ytr[fold == k], ridge_predict(fit, Xtr[fold == k, , drop = FALSE]))
    })))
    a_best <- grid[which.min(cv)]; alphas[i] <- a_best
    preds[i] <- ridge_predict(ridge_fit(Xtr, ytr, a_best), X[i, , drop = FALSE])
  }
  list(preds = preds, alphas = alphas)
}
nested_pcr <- function(X, y, grid, n_inner = 10) {
  n <- length(y); preds <- numeric(n); ks <- integer(n)
  for (i in 1:n) {
    Xtr <- X[-i, ]; ytr <- y[-i]; set.seed(i)
    fold <- sample(rep(1:n_inner, length.out = nrow(Xtr)))
    cv <- sapply(grid, function(k) mean(sapply(1:n_inner, function(f) {
      fit <- pcr_fit(Xtr[fold != f, ], ytr[fold != f], k)
      rmse(ytr[fold == f], pcr_predict(fit, Xtr[fold == f, , drop = FALSE]))
    })))
    k_best <- grid[which.min(cv)]; ks[i] <- k_best
    preds[i] <- pcr_predict(pcr_fit(Xtr, ytr, k_best), X[i, , drop = FALSE])
  }
  list(preds = preds, ks = ks)
}
p_ols_raw <- loocv_ols(X_raw, y_reg)
p_ols_log <- loocv_ols(X_log, y_reg)
nr_raw <- nested_ridge(X_raw, y_reg, alpha_grid)
nr_log <- nested_ridge(X_log, y_reg, alpha_grid)
np_raw <- nested_pcr(X_raw,  y_reg, k_grid)
np_log <- nested_pcr(X_log,  y_reg, k_grid)
```

On log(GDP) the selected $\alpha$ is stable across the 141 outer
folds (range `r sprintf("%.1f–%.1f", min(nr_log$alphas), max(nr_log$alphas))`,
median `r sprintf("%.1f", median(nr_log$alphas))`). On raw GDP it is
less stable (range
`r sprintf("%.2f–%.1f", min(nr_raw$alphas), max(nr_raw$alphas))`,
median `r sprintf("%.1f", median(nr_raw$alphas))`), reflecting a flat
cross-validation surface on the raw scale where different subsets of
140 countries favour very different penalties.

```{r reg-path, fig.cap="Standardised Ridge coefficients as a function of the penalty $\\alpha$, on raw GDP (left) and log GDP (right). The vertical dashed line marks the median $\\alpha$ selected across the 141 nested-LOOCV outer folds.", fig.width=11, fig.height=4}
alphas_path <- 10^seq(-2, 3, length.out = 100)
coef_path <- function(X, y) {
  mx <- colMeans(X); sx <- apply(X, 2, sd)
  Xs <- sweep(sweep(X, 2, mx, "-"), 2, sx, "/"); ys <- y - mean(y)
  XtX <- crossprod(Xs)
  t(sapply(alphas_path, function(a)
    solve(XtX + a * diag(ncol(Xs)), crossprod(Xs, ys))))
}
cp_raw <- coef_path(X_raw, y_reg); cp_log <- coef_path(X_log, y_reg)
years_pred <- seq(1952, 2002, 5)
path_df <- rbind(
  data.frame(alpha = rep(alphas_path, 11),
             coef  = as.vector(cp_raw),
             year  = factor(rep(years_pred, each = length(alphas_path))),
             which = "raw GDP"),
  data.frame(alpha = rep(alphas_path, 11),
             coef  = as.vector(cp_log),
             year  = factor(rep(years_pred, each = length(alphas_path))),
             which = "log GDP"))
marks <- data.frame(which = c("raw GDP","log GDP"),
                    chosen = c(median(nr_raw$alphas), median(nr_log$alphas)))
ggplot(path_df, aes(alpha, coef, colour = year)) +
  geom_line(linewidth = 0.5) +
  geom_vline(data = marks, aes(xintercept = chosen),
             linetype = "dashed", colour = "red", linewidth = 0.5) +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.3) +
  scale_x_log10() + scale_colour_viridis_d(name = "Year") +
  facet_wrap(~which) +
  labs(x = expression(alpha), y = "Standardised coefficient") +
  theme_paper
```

At low $\alpha$ the eleven year coefficients are large and opposing
in sign. This is expected when predictors are near-collinear, since
OLS can produce many combinations of large coefficients that mostly
cancel when predicting the response. Increasing $\alpha$ shrinks the
coefficients toward zero and flattens the year-to-year differences
between them. On log(GDP) the nested-CV median sits in the
intermediate region, where shrinkage has stabilised the coefficients
without yet removing their signal.

```{r reg-table}
k_log <- as.integer(names(sort(table(np_log$ks), decreasing = TRUE))[1])
k_raw <- as.integer(names(sort(table(np_raw$ks), decreasing = TRUE))[1])
results <- data.frame(
  Model = c("OLS (raw GDP)", "OLS (log GDP)",
            "Ridge (raw GDP)", "Ridge (log GDP)",
            sprintf("PCR (raw GDP, k=%d)", k_raw),
            sprintf("PCR (log GDP, k=%d)", k_log)),
  `LOOCV RMSE` = sprintf("%.2f",
    c(rmse(y_reg, p_ols_raw), rmse(y_reg, p_ols_log),
      rmse(y_reg, nr_raw$preds), rmse(y_reg, nr_log$preds),
      rmse(y_reg, np_raw$preds), rmse(y_reg, np_log$preds))),
  `LOOCV R2` = sprintf("%.2f",
    c(r2(y_reg, p_ols_raw), r2(y_reg, p_ols_log),
      r2(y_reg, nr_raw$preds), r2(y_reg, nr_log$preds),
      r2(y_reg, np_raw$preds), r2(y_reg, np_log$preds))),
  check.names = FALSE)
kable(results,
      caption = "Nested leave-one-out cross-validated performance.",
      booktabs = TRUE)
```

OLS on raw GDP gives a LOOCV RMSE of
`r sprintf("%.2f", rmse(y_reg, p_ols_raw))` years and $R^2 =$
`r sprintf("%.2f", r2(y_reg, p_ols_raw))`, while OLS on log(GDP)
gives RMSE `r sprintf("%.2f", rmse(y_reg, p_ols_log))` and
$R^2 =$ `r sprintf("%.2f", r2(y_reg, p_ols_log))`, a
one-and-a-half-year reduction in prediction error from the transform.
Ridge on log(GDP) improves this further to RMSE
`r sprintf("%.2f", rmse(y_reg, nr_log$preds))` and $R^2 =$
`r sprintf("%.2f", r2(y_reg, nr_log$preds))`, while Ridge on raw GDP
reaches RMSE `r sprintf("%.2f", rmse(y_reg, nr_raw$preds))` and
$R^2 =$ `r sprintf("%.2f", r2(y_reg, nr_raw$preds))`, a modest gain
over the raw-scale OLS. GDP's skewness makes the linear relationship
with life expectancy non-linear on the raw scale but close to linear
on the log scale, and the regression results express the cost of
ignoring this as 0.19 units of $R^2$.

PCR has the lowest RMSE of the six models. On log(GDP) the
inner CV selected $k = 3$ components in every one of the 141 outer
folds, and the resulting nested LOOCV RMSE is
`r sprintf("%.2f", rmse(y_reg, np_log$preds))` with $R^2 =$
`r sprintf("%.2f", r2(y_reg, np_log$preds))`. This stability is
diagnostic of the underlying structure. The PCA of the GDP block
showed that the first three components already explain 99% of the
variance, so removing any single country never changes which
directions dominate the design matrix and the inner CV reaches the
same conclusion every time. PCR on raw GDP peaks at $k = 2$ (modal
selection) with RMSE `r sprintf("%.2f", rmse(y_reg, np_raw$preds))`,
but still does not reach the log-based models, confirming that the
transform is more important than the multicollinearity fix.

The nested design lets us separate two effects. For PCR the held-out
country never influences the $k$ used to score it (because $k$ is
constant across folds), so the nested RMSE is an unbiased estimate of
out-of-sample RMSE. For Ridge the inner-selected $\alpha$ does vary
across outer folds, so the nested score exceeds what a single flat
LOOCV score would suggest. PCR's zero tuning variance makes it the
preferred model of the two. Both outperform OLS on log(GDP) because
they address the near-collinearity of the year columns, with PCR
doing so more cleanly.

```{r reg-best, fig.cap="Leave-one-out predictions against actual 2007 life expectancy for the best model.", fig.width=7, fig.height=6}
best_idx <- which.min(c(rmse(y_reg, p_ols_raw), rmse(y_reg, p_ols_log),
                        rmse(y_reg, nr_raw$preds), rmse(y_reg, nr_log$preds),
                        rmse(y_reg, np_raw$preds), rmse(y_reg, np_log$preds)))
best_preds <- list(p_ols_raw, p_ols_log, nr_raw$preds, nr_log$preds,
                   np_raw$preds, np_log$preds)[[best_idx]]
pd <- data.frame(actual = y_reg, predicted = best_preds,
                 country = country, continent = continent)
pd$ra <- abs(pd$predicted - pd$actual)
worst3 <- pd[order(-pd$ra), ][1:3, ]
ggplot(pd, aes(actual, predicted, colour = continent)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 2, alpha = 0.75) +
  geom_text_repel(data = worst3, aes(label = country),
                  size = 3, fontface = "bold", show.legend = FALSE) +
  scale_colour_manual(values = palette_continents) +
  labs(x = "Actual 2007 life expectancy (years)",
       y = "LOOCV predicted life expectancy (years)",
       title = results$Model[best_idx], colour = "Continent") +
  theme_paper
```

```{r reg-residuals, include=FALSE}
resid_all <- best_preds - y_reg
sd_r <- sd(resid_all); thr <- 2 * sd_r
outliers <- which(abs(resid_all) > thr)
```

The residuals have a clear structure. The standard deviation of the
141 leave-one-out residuals is `r sprintf("%.2f", sd_r)`, and
`r length(outliers)` countries sit beyond two standard deviations of
the fit. All seven are African and all are over-predicted. Swaziland,
Botswana, and South Africa lie above the diagonal by more than 20
years, with Angola, Lesotho, Equatorial Guinea, and Gabon between 15
and 17 years. Zambia, Namibia, and Nigeria sit just inside the two-SD
threshold, extending the same pattern. The direction is systematic
rather than random. Countries with pre-1990 GDP histories comparable
to other middle-income nations experienced two shocks the model
cannot see from GDP alone. The first is a health cluster in
southern Africa (Swaziland, Botswana, Lesotho, South Africa, Zambia,
Namibia), where life expectancy fell between 1990 and 2007 without a
proportional fall in income. The second is an
oil-rent pattern (Equatorial Guinea, Gabon, Angola), where per-capita
GDP was pushed up by resource extraction concentrated in a narrow
elite without matching gains in public health. The two
negative-residual outliers are Vietnam ($-14.2$ years, life expectancy
outpaces GDP) and Nicaragua ($-13.0$), which show the opposite gap. A
health shock that leaves GDP largely unaffected, or a health gain
decoupled from GDP, is invisible to the model, and these countries
sit furthest from the diagonal under every combination of transform,
regularisation, and component count we tried.
