---
title: "MATH3030 Coursework"
subtitle: "Multivariate Statistics --- Spring 2026"
author: "Aayush Parekh (20527121)"
date: ""
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: pdflatex
    fig_caption: yes
fontsize: 11pt
geometry: margin=2.5cm
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{microtype}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  dpi        = 200,
  cache      = TRUE,
  fig.pos    = "H",
  out.width  = "100%"
)
user_lib <- file.path(path.expand("~"), "R", "library")
.libPaths(c(user_lib, .libPaths()))
suppressPackageStartupMessages({
  library(ggplot2); library(ggrepel); library(MASS)
  library(mclust); library(patchwork); library(scales); library(knitr)
})
set.seed(42)

palette_continents <- c(
  "Africa"="#66c2a5","Americas"="#fc8d62","Asia"="#8da0cb",
  "Europe"="#e78ac3","Oceania"="#a6d854"
)
theme_paper <- theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 10))

UN <- read.csv("UN.csv")
years       <- seq(1952, 2007, 5)
year_labels <- as.character(years)
continent   <- UN[, 1]
country     <- UN[, 2]
gdp     <- UN[,  3:14]; colnames(gdp)     <- year_labels; rownames(gdp)     <- country
lifeExp <- UN[, 15:26]; colnames(lifeExp) <- year_labels; rownames(lifeExp) <- country
popn    <- UN[, 27:38]; colnames(popn)    <- year_labels; rownames(popn)    <- country
```

\newpage

# Question 1

## Exploratory Data Analysis

The UN dataset contains GDP per capita, life expectancy, and population
for 141 countries at 12 evenly spaced time points from 1952 to 2007.
No values are missing. Sample sizes per continent are uneven (52
African, 32 Asian, 30 European, 25 American, and 2 Oceanian), so any
continent-level summary for Oceania is effectively a two-country
average of Australia and New Zealand. We summarise the 70-year
dynamics with continent-mean trajectories for each variable and a
table of 1952 and 2007 medians.

```{r eda-trends, fig.cap="Continent-mean trajectories, 1952–2007. GDP and population are on log scales; life expectancy is linear.", fig.width=10, fig.height=3.2}
eda_long <- function(data, varname) {
  mat <- as.matrix(data)
  do.call(rbind, lapply(sort(unique(continent)), function(cont) {
    m_year <- colMeans(mat[continent == cont, , drop = FALSE])
    data.frame(year = years, continent = cont, value = as.numeric(m_year),
               variable = varname, stringsAsFactors = FALSE)
  }))
}
eda_trends <- rbind(
  eda_long(gdp,     "GDP per capita (USD)"),
  eda_long(lifeExp, "Life expectancy (years)"),
  eda_long(popn,    "Population")
)
panels <- lapply(c("GDP per capita (USD)","Life expectancy (years)","Population"),
  function(v) {
    d <- subset(eda_trends, variable == v)
    p <- ggplot(d, aes(year, value, colour = continent)) +
      geom_line(linewidth = 0.9) + geom_point(size = 1.3) +
      scale_colour_manual(values = palette_continents) +
      labs(x = "Year", y = NULL, title = v) + theme_paper
    if (v != "Life expectancy (years)") p <- p + scale_y_log10(labels = label_comma())
    else p <- p + scale_y_continuous(labels = label_comma())
    p
  })
wrap_plots(panels, ncol = 3, guides = "collect") &
  theme(legend.position = "right") &
  labs(colour = "Continent")
```

Asia gains most on both axes. Mean life expectancy rose from 44 to 72
years and median GDP per capita almost quadrupled. The continent mean
hides a wide spread. Singapore, South Korea, and Taiwan each grew
GDP per capita by 20–24×, far above the 3.8× Asian median. China
alone accounts for the population panel, rising from 556 million in
1952 to 1.32 billion in 2007.

Africa's trajectory is much shallower. Median GDP rose only 1.5× over
55 years, and mean life expectancy, after rising from 39 to 53 years,
plateaued from the 1990s onward. The plateau coincides with a broad
health and conflict downturn in southern and central Africa. Zimbabwe's life expectancy fell from 48.5 to 43.5 years
(the largest decline in the dataset), Swaziland, Zambia, and Lesotho
show similar stagnation, and the Democratic Republic of the Congo is
the only country whose GDP per capita ended below its 1952 level, at
36% of its original value. Africa's population grew faster than any
other continent's, which compounds the GDP-growth shortfall on a
per-capita basis.

Europe and Oceania remain at the top throughout. Europe's GDP growth
(\$5,143 to \$28,054, 5.5×) matches Asia's in multiplier terms, but
its life-expectancy gain of +13 years is about half of Asia's +28.
This is a ceiling effect. A population already close to the
demographic ceiling has less room to gain years than one catching up.
Both continents evolve more smoothly than Asia or Africa, with no
sharp deviations between adjacent five-year intervals.

```{r eda-bubble, fig.cap="GDP per capita vs life expectancy in 2007. Bubble size is proportional to population.", fig.width=7, fig.height=4.8}
bubble_df <- data.frame(country = country, continent = continent,
                        gdp = gdp[,"2007"], lifeExp = lifeExp[,"2007"],
                        popn = popn[,"2007"])
largest <- do.call(rbind, by(bubble_df, bubble_df$continent,
  function(d) d[which.max(d$popn), ]))
ggplot(bubble_df, aes(gdp, lifeExp, size = popn, colour = continent)) +
  geom_point(alpha = 0.6) +
  geom_text_repel(data = largest, aes(label = country),
                  size = 3, fontface = "bold", show.legend = FALSE) +
  scale_x_log10(labels = label_comma()) +
  scale_size_area(max_size = 15, guide = "none") +
  scale_colour_manual(values = palette_continents) +
  labs(x = "GDP per capita (log scale, 2007)",
       y = "Life expectancy (years, 2007)", colour = "Continent") +
  theme_paper
```

GDP and life expectancy correlate at 0.81 on the log scale and 0.68
on the raw, the gap reflecting diminishing returns on life expectancy
at high income. Asia is the only continent spanning the full income
range, from India and Bangladesh up to Japan and Singapore, with
China between the two modes.

```{r eda-kde, fig.cap="Kernel density estimates of each variable in 2007.", fig.width=11, fig.height=3.2}
kde_specs <- list(
  list(name="log10(GDP per capita, 2007)", x=log10(gdp[,"2007"])),
  list(name="Life expectancy (2007)",      x=lifeExp[,"2007"]),
  list(name="log10(Population, 2007)",     x=log10(popn[,"2007"])))
kde_panels <- lapply(kde_specs, function(s) {
  d <- data.frame(x = s$x, continent = continent)
  ggplot(d, aes(x, fill = continent, colour = continent)) +
    geom_density(alpha = 0.2, linewidth = 0.7) +
    scale_fill_manual(values = palette_continents, guide = "none") +
    scale_colour_manual(values = palette_continents) +
    labs(x = NULL, y = "Density", title = s$name, colour = "Continent") +
    theme_paper + theme(axis.text.y = element_blank(),
                        axis.ticks.y = element_blank())
})
wrap_plots(kde_panels, ncol = 3, guides = "collect") &
  theme(legend.position = "right")
```

Asia's GDP distribution is bimodal, splitting Gulf states and East
Asian tigers from South and Central Asia, with an IQR ratio of 8.9,
the largest of any continent. Africa's life-expectancy distribution
is wide and left-skewed, with a sub-45-year tail in southern African
countries. Asia carries the long population right tail (China, India).
The continent means hide all of this.

```{r eda-table}
summary_tbl <- data.frame(
  N          = as.vector(table(continent)),
  `GDP 1952` = round(tapply(gdp[,"1952"],     continent, median), 0),
  `GDP 2007` = round(tapply(gdp[,"2007"],     continent, median), 0),
  `LE 1952`  = round(tapply(lifeExp[,"1952"], continent, median), 1),
  `LE 2007`  = round(tapply(lifeExp[,"2007"], continent, median), 1),
  `Pop 2007 (M)` = round(tapply(popn[,"2007"], continent, sum) / 1e6, 0),
  check.names = FALSE
)
rownames(summary_tbl) <- sort(unique(continent))
kable(summary_tbl,
      caption = "Median GDP per capita (USD) and life expectancy (years) by continent; population (millions).",
      booktabs = TRUE)
```

```{r eda-change, fig.cap="GDP-per-capita multiplier (1952$\\rightarrow$2007, log scale) against absolute change in life expectancy. Labelled points are the largest deviations on either axis.", fig.width=7, fig.height=4.5}
change_df <- data.frame(
  country   = country,
  continent = continent,
  gdp_mult  = gdp[, "2007"] / gdp[, "1952"],
  le_diff   = lifeExp[, "2007"] - lifeExp[, "1952"]
)
ext_idx <- unique(c(
  order(change_df$gdp_mult, decreasing = TRUE)[1:5],
  order(change_df$gdp_mult)[1:3],
  order(change_df$le_diff)[1:5],
  order(change_df$le_diff, decreasing = TRUE)[1:3]
))
ggplot(change_df, aes(gdp_mult, le_diff, colour = continent)) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
  geom_vline(xintercept = 1, colour = "grey60", linewidth = 0.3) +
  geom_point(size = 1.8, alpha = 0.75) +
  geom_text_repel(data = change_df[ext_idx, ], aes(label = country),
                  size = 2.5, show.legend = FALSE, max.overlaps = Inf) +
  scale_x_log10() +
  scale_colour_manual(values = palette_continents) +
  labs(x = "GDP per capita multiplier, 1952–2007 (log scale)",
       y = "Change in life expectancy (years)",
       colour = "Continent") +
  theme_paper
```

A small number of countries deviate strongly from their continent's
trend and reappear as extreme points in the PCA, CCA, and MDS
projections below. Equatorial Guinea's GDP per capita rose
32× on the back of offshore oil discoveries, Zimbabwe's life
expectancy fell by five years, Singapore, South Korea, and Taiwan
achieved sustained 20–24× GDP multiples, and the Democratic Republic
of the Congo and Haiti are the two economies that regressed most
sharply, ending the period at 36% and 65% of their 1952 GDP per
capita respectively. The two axes are largely decoupled, with top
GDP multipliers and top life-expectancy gains landing on different
countries.

## Principal Component Analysis

We perform PCA independently on the $141 \times 12$ GDP and
life-expectancy blocks. GDP values span more than two orders of
magnitude across countries, so we standardise each year column (which
is equivalent to decomposing the correlation matrix) to prevent
larger recent-year variances from dominating. Life expectancy is on a
comparable scale across years, but we also standardise it. The
covariance and correlation decompositions produce near-identical
results (PC1 explains 92.0% and 92.3% of the variance respectively),
and standardising keeps the treatment of the two variable blocks
consistent.

```{r pca-fit}
pca_gdp  <- prcomp(gdp,     scale. = TRUE)
pca_life <- prcomp(lifeExp, scale. = TRUE)
var_ratio <- function(p) (p$sdev^2) / sum(p$sdev^2)
```

### Scree analysis and variance explained

```{r pca-scree, fig.cap="Scree plots for the GDP and life-expectancy PCAs.", fig.width=11, fig.height=4}
scree_df <- function(p, label) data.frame(pc = seq_along(var_ratio(p)),
  var = var_ratio(p), cum = cumsum(var_ratio(p)), which = label)
scree <- rbind(scree_df(pca_gdp, "GDP per capita"),
               scree_df(pca_life, "Life expectancy"))
ggplot(scree, aes(pc)) +
  geom_col(aes(y = var), fill = "steelblue", alpha = 0.6) +
  geom_line(aes(y = cum), colour = "darkred") +
  geom_point(aes(y = cum), colour = "darkred", size = 1.3) +
  geom_hline(yintercept = 0.9, linetype = "dashed",
             colour = "grey40", linewidth = 0.3) +
  facet_wrap(~which) +
  scale_x_continuous(breaks = 1:12) +
  labs(x = "Principal component", y = "Proportion of variance") +
  theme_paper
```

Both decompositions concentrate almost all variance in a single
component. PC1 explains 91.8% of the GDP variance and 92.3% of the
life-expectancy variance while PC2 adds a further 5.3% and 5.6%. Two
components cover 97.1% of GDP and 97.9% of life expectancy. Adjacent
years correlate at $\approx 0.99$, so 12 columns compress to two.

### Interpretation of PC1 and PC2

The loadings have the same structure on both decompositions. PC1
assigns nearly uniform positive weights across all 12 years (0.28 to
0.30 on GDP, 0.27 to 0.30 on life expectancy), so a country's PC1
score is approximately proportional to its average level over the
full 55-year period. For PC2, the loadings begin strongly negative in
1952 ($-0.35$) and rise monotonically to strongly positive in 2007
($+0.42$). A positive PC2 score therefore indicates a country whose
values were low early and high late (a catching-up or fast-growth
trajectory), and a negative PC2 score indicates the opposite
(stagnation or decline from a relatively high base).

### PC score scatter plots

```{r pca-align, include=FALSE}
sw <- which(country == "Switzerland")
if (length(sw) && pca_gdp$x[sw, 1] < 0)  pca_gdp$x[, 1]  <- -pca_gdp$x[, 1]
if (length(sw) && pca_life$x[sw, 1] < 0) pca_life$x[, 1] <- -pca_life$x[, 1]
sg <- which(country == "Singapore")
if (length(sg) && pca_gdp$x[sg, 2] < 0) pca_gdp$x[, 2] <- -pca_gdp$x[, 2]
om <- which(country == "Oman")
if (length(om) && pca_life$x[om, 2] < 0) pca_life$x[, 2] <- -pca_life$x[, 2]
```

```{r pca-scores, fig.cap="PC1 vs PC2 scatter plots for GDP (left) and life expectancy (right), coloured by continent.", fig.width=13, fig.height=5}
scatter_df <- function(p, label) {
  d <- data.frame(PC1 = p$x[,1], PC2 = p$x[,2], country = country,
                  continent = continent, which = label)
  d$dist <- sqrt(d$PC1^2 + d$PC2^2); d
}
make_scatter <- function(d, title) {
  top <- d[order(-d$dist), ][1:8, ]
  ggplot(d, aes(PC1, PC2, colour = continent)) +
    geom_point(size = 1.8, alpha = 0.75) +
    geom_text_repel(data = top, aes(label = country),
                    size = 2.3, show.legend = FALSE, max.overlaps = Inf) +
    scale_colour_manual(values = palette_continents) +
    labs(title = title, colour = "Continent") + theme_paper
}
(make_scatter(scatter_df(pca_gdp,  "GDP"), "GDP — PC1 vs PC2") |
 make_scatter(scatter_df(pca_life, "Life Expectancy"), "Life Expectancy — PC1 vs PC2")) +
  plot_layout(guides = "collect") & theme(legend.position = "right")
```

On the GDP scatter PC1 ranks countries by income level (Switzerland
and the US at the top, the poorest African economies at the bottom).
PC2 then picks out growth trajectory, with the East Asian tigers
clustering at the top, while Venezuela, Iraq, and Libya sit at the
negative end as 1952 high-income economies that lost ground.

The life-expectancy scatter has the same shape. PC1 again ranks by
level (Scandinavian and Alpine Europe at the top, sub-Saharan Africa
at the bottom) and PC2 by trajectory, with Oman and Vietnam at one
end and Zimbabwe and Botswana at the other, matching the southern
African downturn from the EDA. Asia stretches along PC2 because of
its bimodal income distribution; Europe and Oceania sit together at
the top of PC1, Africa at the bottom, the Americas between.

### GDP PC1 versus life-expectancy PC1

```{r pca-cross, fig.cap="First PC scores, GDP against life expectancy.", fig.width=7, fig.height=5.5}
cross <- data.frame(gdp_pc1 = pca_gdp$x[,1], life_pc1 = pca_life$x[,1],
                    country = country, continent = continent)
cross$dist <- sqrt(cross$gdp_pc1^2 + cross$life_pc1^2)
ggplot(cross, aes(gdp_pc1, life_pc1, colour = continent)) +
  geom_point(size = 2, alpha = 0.75) +
  geom_text_repel(data = cross[order(-cross$dist), ][1:10, ],
                  aes(label = country), size = 2.4, show.legend = FALSE) +
  scale_colour_manual(values = palette_continents) +
  labs(x = "GDP PC1 score", y = "Life Expectancy PC1 score", colour = "Continent") +
  theme_paper
```

The two PC1 scores correlate at
`r round(cor(pca_gdp$x[,1], pca_life$x[,1]), 2)`, in line with the
2007 cross-sectional 0.81 on log(GDP). Most countries lie close to
the diagonal; deviations are countries whose health outcomes are
stronger or weaker than income alone would predict, picking up
public-health investment, diet, and health-system access.

### Limitations of linear PCA

The decomposition is restricted to linear combinations of the year
columns and the GDP–life expectancy relationship is non-linear
(Pearson correlation 0.68 on raw GDP against 0.81 on log(GDP)).
Kernel PCA, which applies the standard decomposition in a feature
space induced by a kernel, would capture this curvature directly and
would be expected to produce tighter continent separation than the
linear projection used here.^[Hastie, Tibshirani, and Friedman,
*The Elements of Statistical Learning* (2nd ed., 2009), Chapter 14.]

## Canonical Correlation Analysis

We apply CCA with $\log(\text{GDP})$ as the first variable set and
life expectancy as the second, giving $p = q = 12$. The log transform
is motivated by the same skewness and non-linearity noted in the EDA
and the PCA limitations subsection. Raw GDP is strongly right-skewed
(skewness 1.88) and the 2007 GDP–life expectancy relationship curves
noticeably on the raw scale but is close to linear on the log scale.

```{r cca-fit}
log_gdp  <- log(as.matrix(gdp))
life_mat <- as.matrix(lifeExp)
cc  <- cancor(log_gdp, life_mat)
cc_raw <- cancor(as.matrix(gdp), life_mat)
U <- scale(log_gdp,  scale = FALSE) %*% cc$xcoef
V <- scale(life_mat, scale = FALSE) %*% cc$ycoef
```

```{r cca-scatter, fig.cap="First pair of canonical variables, coloured by continent.", fig.width=7, fig.height=5.5}
cca_df <- data.frame(U1 = U[,1], V1 = V[,1],
                     country = country, continent = continent)
cca_df$dist <- sqrt(cca_df$U1^2 + cca_df$V1^2)
ggplot(cca_df, aes(U1, V1, colour = continent)) +
  geom_point(size = 2, alpha = 0.75) +
  geom_text_repel(data = cca_df[order(-cca_df$dist), ][1:8, ],
                  aes(label = country), size = 2.4, show.legend = FALSE) +
  scale_colour_manual(values = palette_continents) +
  labs(x = "First CC variable (log GDP side)",
       y = "First CC variable (life expectancy side)",
       title = sprintf("CCA: First canonical pair  (r = %.3f)", cc$cor[1]),
       colour = "Continent") +
  theme_paper
```

### Canonical correlations

The first canonical correlation is `r round(cc$cor[1], 2)`, and
subsequent correlations decay to `r round(cc$cor[2], 2)`,
`r round(cc$cor[3], 2)`, and `r round(cc$cor[4], 2)`, with the
remainder below 0.40. The first pair therefore carries most of the
shared linear structure between the two blocks. Looking at CC2
directly, the GDP-side loadings start near zero in 1952 (0.003), rise
to 0.11 around 1977, and fall back to $-0.11$ by 2007, while the
life-expectancy loadings are uniformly small and negative across all
12 years ($-0.05$ to $-0.19$). The GDP pattern alone describes a
faint mid-period growth contrast, but it is not matched by any
equivalent temporal structure on the life-expectancy side, so the two
variates of CC2 are not aligned on the same temporal feature and the
pair has no joint interpretation. Higher-order pairs decay further
and are similarly uninterpretable.

### Interpretation of the first canonical pair

The first canonical axis ranks countries by a joint development
level that log(GDP) and life expectancy identify together. The
ordering is near-identical to PCA's on either variable alone, because
the 12 year columns within each block correlate at $\approx 0.99$ and
CCA has little room to pick a direction meaningfully different from
PCA's. The continent structure reproduces the PCA pattern, with
Africa lower-left and Europe and Oceania upper-right.

### Effect of the log transform

On raw GDP the first canonical correlation drops to
`r round(cc_raw$cor[1], 2)` from `r round(cc$cor[1], 2)`. CCA only
finds linear combinations, so curvature in the raw GDP–life
expectancy relationship is uncaptured; the log transform straightens
it and recovers the lost correlation.

## Multidimensional Scaling

Classical (metric) MDS is applied to the combined 36-dimensional
dataset formed by concatenating log(GDP), life expectancy, and
log(population), using Euclidean distance as specified in the
coursework. The log transforms on GDP and population bring them onto
a roughly comparable scale with life expectancy.

```{r mds-fit}
UN_transformed <- cbind(log(as.matrix(gdp)), as.matrix(lifeExp), log(as.matrix(popn)))
d_full    <- dist(UN_transformed)
mds       <- cmdscale(d_full, k = 2, eig = TRUE)
d_reduced <- dist(mds$points)
dist_cor  <- cor(as.vector(d_full), as.vector(d_reduced))
```

The Pearson correlation between pairwise distances in the original
36-D space and the 2D MDS projection is
`r round(dist_cor, 3)`, so the two retained dimensions preserve the
distance geometry of the full data almost exactly.

```{r mds-plot, fig.cap="Two-dimensional MDS representation, coloured by continent.", fig.width=8, fig.height=6}
mds_df <- data.frame(D1 = mds$points[,1], D2 = mds$points[,2],
                     country = country, continent = continent)
mds_df$dist <- sqrt(mds_df$D1^2 + mds_df$D2^2)
ggplot(mds_df, aes(D1, D2, colour = continent)) +
  geom_point(size = 2, alpha = 0.75) +
  geom_text_repel(data = mds_df[order(-mds_df$dist), ][1:10, ],
                  aes(label = country), size = 2.4, show.legend = FALSE) +
  scale_colour_manual(values = palette_continents) +
  labs(x = "MDS dimension 1", y = "MDS dimension 2", colour = "Continent") +
  theme_paper
```

### Development axis

Dimension 1 is a development axis. The continent centroids line up
from Africa at $-37$ through the Americas at $+18$ and Europe at
$+44$ to Oceania at $+52$, the same order PCA and CCA produced. The
country ordering inside is near-identical too, since all three
methods project the same dominant development signal.

### Growth direction

Dimension 2 separates countries by trajectory rather than level,
with gainers like Oman and Vietnam at one end and southern African
declines (Zimbabwe, Botswana) at the other. Asia's centroid sits below the
horizontal because its fast-gaining countries dominate; Africa's
hovers near zero because the continent contains both improving and
declining cases. This is the same contrast as PC2 in the
life-expectancy PCA, now found jointly with all three variables.

### Relationship to PCA

The agreement between the MDS and PCA axes is not a coincidence.
Classical MDS with Euclidean distance is, up to sign, the same
decomposition as PCA on the same centered data. Writing the SVD of
the centered data matrix as
$$X = U D V^T,$$
PCA returns the scores
$$T_{\text{PCA}} = X V = U D.$$
Classical MDS forms the Gram matrix $B = X X^T$ and takes the
eigendecomposition
$$B = U D^2 U^T,$$
from which the MDS coordinates are the leading columns of $U D$, the
same matrix as $T_{\text{PCA}}$. The MDS scatter does not literally
reproduce either PCA scatter because the input differs. Here MDS is
applied to the 36-column concatenation of $\log(\text{GDP})$, life
expectancy, and $\log(\text{popn})$, whereas the PCA sections ran the
decomposition on the GDP and life-expectancy blocks separately.

### Comparison with the earlier analyses

MDS recovers the same two axes the PCAs found, but with all three
variables at once instead of one block at a time. The 0.997 distance
correlation says development level and growth direction are the two
dominant sources of variation in the full UN data.

\newpage

# Question 2

## Linear Discriminant Analysis

We use LDA to predict continent from all 36 features (GDP, life
expectancy, and population across 12 years). The data are split 70/30
into training and test sets.

```{r lda-fit}
X_all <- as.matrix(cbind(gdp, lifeExp, popn))
y_all <- factor(continent)
set.seed(42)
n <- nrow(X_all)
idx <- sample(n, round(0.7 * n))
fit_lda <- MASS::lda(X_all[idx, ], grouping = y_all[idx])
pred    <- predict(fit_lda, X_all[-idx, ])$class
acc_single <- mean(pred == y_all[-idx])
cm <- table(True = y_all[-idx], Pred = pred)

accs <- numeric(50)
for (s in 1:50) {
  set.seed(s)
  i_tr <- sample(n, round(0.7 * n))
  m    <- MASS::lda(X_all[i_tr, ], grouping = y_all[i_tr])
  p    <- predict(m, X_all[-i_tr, ])$class
  accs[s] <- mean(p == y_all[-i_tr])
}
```

A single fixed split with seed 42 gives a test accuracy of
`r sprintf("%.3f", acc_single)` on `r nrow(X_all) - length(idx)`
held-out countries, and averaging over 50 random splits gives a more
stable estimate of `r sprintf("%.3f ± %.3f", mean(accs), sd(accs))`.
Both figures sit comfortably above the 0.20 five-class chance rate
and above the 0.37 majority-class baseline set by Africa's sample
proportion, so the classifier is extracting genuine structure from
the features.

```{r lda-confusion}
kable(as.matrix(cm),
      caption = "Rows are true continents, columns are predicted.",
      booktabs = TRUE)
```

Per-class accuracy varies substantially. Africa is the best-classified
continent, followed by Europe, Asia, and the Americas. The Oceanian
country in the test set is misclassified as European. Asia's errors
are spread across Africa, the Americas, and Europe in roughly equal
proportions.

The per-class pattern follows from LDA's assumptions. The rule
$$\delta_c(x) = x^\top \Sigma^{-1} \mu_c - \tfrac{1}{2} \mu_c^\top \Sigma^{-1} \mu_c + \log \pi_c,$$
with $\mu_c$ the class mean, $\Sigma$ the pooled within-class
covariance, and $\pi_c$ the prior, treats each class as Gaussian with
shared $\Sigma$. Asia breaks the first assumption. Its GDP
distribution is bimodal, so $\mu_{\text{Asia}}$ lands between the two
modes and $N(\mu_{\text{Asia}}, \Sigma)$ concentrates where almost no
Asian country actually sits, while overlapping the Gaussians for
Africa, the Americas, and Europe. Oceania and Europe have far
tighter within-class spread than Africa or Asia, so the pooled
$\Sigma$ fits neither. Africa is classified well despite being
heterogeneous because its spread runs along a single unimodal axis
the Gaussian can still represent.

## Clustering

Three methods from the module are applied to the standardised GDP and
life-expectancy blocks (24 features). k-means and the model-based
clustering are fitted directly, and hierarchical clustering is
applied via its linkage-distance dendrogram, using Ward's linkage as
in the computer lab sessions. Ward merges the pair of clusters
producing the smallest increase in within-cluster variance at each
step. $k$ is chosen from the dendrogram and the MVN-mixture BIC
curve, with the same $k$ then used for k-means so that all three
methods can be compared on equal footing.

```{r clust-fit}
X_cluster <- scale(cbind(as.matrix(gdp), as.matrix(lifeExp)))
h         <- hclust(dist(X_cluster), method = "ward.D2")
mc_models <- Mclust(X_cluster, G = 1:8, verbose = FALSE)
bic_per_G <- apply(mc_models$BIC, 1, function(v) max(v, na.rm = TRUE))
```

```{r clust-selection, fig.cap="Hierarchical clustering dendrogram (left) and model-based BIC as a function of number of components (right).", fig.width=11, fig.height=4}
par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(h, labels = FALSE, main = "Hierarchical clustering (Ward)",
     xlab = "Countries", sub = "", ylab = "Linkage distance")
plot(as.numeric(names(bic_per_G)), bic_per_G, type = "b",
     xlab = "Number of components G", ylab = "BIC (best model)",
     main = "Model-based (mclust) BIC", col = "steelblue", pch = 19)
```

### Choice of k

The dendrogram's three largest merge distances are
`r sprintf("%.1f, %.1f, and %.1f", sort(h$height, decreasing=TRUE)[1], sort(h$height, decreasing=TRUE)[2], sort(h$height, decreasing=TRUE)[3])`,
with every subsequent merge below 16. The
`r sprintf("%.0f", sort(h$height, decreasing=TRUE)[3] - sort(h$height, decreasing=TRUE)[4])`-unit
gap between merges 3 and 4 identifies $k = 3$ as the cleanest cut.
Mclust's BIC peaks at $G = 2$ (`r round(bic_per_G[2], 0)`), with
$G = 3$ second at `r round(bic_per_G[3], 0)` and $G = 4$ dropping
to `r round(bic_per_G[4], 0)`. The
`r round(bic_per_G[2] - bic_per_G[3], 0)`-unit gap from $G = 2$ to
$G = 3$ is small next to the
`r round(bic_per_G[3] - bic_per_G[4], 0)`-unit drop to $G = 4$, so
the BIC profile leaves $G = 2$ and $G = 3$ the only credible
candidates and rules out the rest.

The two criteria measure different things. The dendrogram tracks
where merges introduce new within-cluster spread. BIC scores
per-parameter fit of a Gaussian mixture density model. PCA and MDS
in Question 1 showed the data run along a single development axis,
and a 2-component Gaussian fits that shape by placing one mode at
each tail. A third component then costs more parameters than it
gains in likelihood, so BIC prefers $G = 2$ even when the data along
the axis split into three groups.

We use $k = 3$. The dendrogram has a clean three-merge cut, PCA and
MDS already showed a low–middle–high ordering, and the $G = 3$
mixture is the second-best BIC fit. A two-cluster cut collapses the
middle group into the low one.

### Agreement between methods

```{r clust-compare}
best_k <- 3
km_res   <- kmeans(X_cluster, centers = best_k, nstart = 10)
hier_res <- cutree(h, k = best_k)
mc_res   <- Mclust(X_cluster, G = best_k, verbose = FALSE)$classification
ari_km_h  <- adjustedRandIndex(km_res$cluster, hier_res)
ari_km_mc <- adjustedRandIndex(km_res$cluster, mc_res)
ari_h_mc  <- adjustedRandIndex(hier_res,       mc_res)
```

```{r clust-comparison, fig.cap="Cluster assignments at k = 3 for k-means, hierarchical clustering, and the MVN mixture.", fig.width=13, fig.height=4}
pca2 <- prcomp(X_cluster)$x[, 1:2]
cdf <- data.frame(PC1 = pca2[,1], PC2 = pca2[,2],
                  kmeans = factor(km_res$cluster),
                  hier   = factor(hier_res),
                  mvn    = factor(mc_res),
                  continent = continent)
pl <- function(lab, ti) ggplot(cdf, aes(PC1, PC2, colour = .data[[lab]])) +
  geom_point(size = 1.8, alpha = 0.8) + labs(title = ti, colour = lab) +
  theme_paper
(pl("kmeans", "k-means (k=3)") | pl("hier", "Hierarchical (k=3)") |
 pl("mvn", "MVN mixture (G=3)")) &
  theme(legend.position = "none")
```

At $k = 3$ the Adjusted Rand Index is `r round(ari_km_h, 2)` between
k-means and hierarchical clustering, `r round(ari_km_mc, 2)` between
k-means and the MVN mixture, and `r round(ari_h_mc, 2)` between
hierarchical and the MVN mixture. The k-means and hierarchical
results are nearly identical, consistent with both being
variance-minimising methods on the same standardised features. The
MVN mixture disagrees more because mclust fits full-rank covariance
structures (varying shape, equal volume) rather than the spherical
within-cluster geometry that k-means and Ward's method assume, and
the covariance flexibility lets it draw cluster boundaries that
follow the anisotropic spread of the data rather than minimising
within-cluster variance.

### Cluster interpretation

The clusters split by development level. One holds 18 of 30 European
countries, 6 Asian, and both Oceanian, sitting at the high end. A
second holds 42 of 52 African countries with 10 Asian and 2 American,
at the low end. The third is the middle group, dominated by the
Americas (21 of 25) with 16 Asian and the remaining 12 European
countries. Continent is a reasonable but imperfect proxy. Africa and
Europe fall almost entirely into the low and high clusters, while the
Americas and Asia split across the middle and the extremes.

## Linear Regression

We predict 2007 life expectancy from the 11 GDP columns for 1952–2002,
excluding 2007 GDP to avoid contemporaneous leakage. The 11
predictors correlate at $\approx 0.99$, so OLS alone is a poor
baseline; we compare it to ridge regression and principal component
regression (PCR), each fitted on raw GDP and on $\log(\text{GDP})$.

Performance is measured by leave-one-out cross-validation. Each
country is held out in turn, the model is refit on the remaining 140,
and the held-out country is predicted from that fit.^[Hastie,
Tibshirani, and Friedman, *The Elements of Statistical Learning*
(2nd ed., 2009), Section 7.10.] The score is
$$\text{RMSE}_{\text{LOOCV}} = \sqrt{\tfrac{1}{141} \sum_{i=1}^{141} (y_i - \hat y_i)^2},$$
where each $\hat y_i$ comes from a fit that never saw country $i$.

Ridge and PCR have hyperparameters ($\alpha$ and $k$), which we tune
by nested cross-validation. The LOOCV above is the outer loop. Within
each outer fold, an inner 10-fold CV on the 140 training rows picks
$\alpha$ or $k$, and the selected value is used to refit on those 140
rows and predict the held-out country. The held-out country never
enters the inner loop, so it cannot influence the hyperparameter used
to score it. A single flat LOOCV that chose $\alpha$ or $k$ from all
141 residuals would feed each country's residual back into its own
hyperparameter selection, pulling the reported RMSE below the true
out-of-sample value. OLS has no hyperparameter and is unaffected.

### Ridge regression and principal component regression

Ridge regression minimises
$$\|y - X\beta\|^2 + \alpha \|\beta\|^2,$$
adding a penalty on the squared length of $\beta$ to the usual
least-squares loss. When predictors are near-collinear, OLS fits tend
to pair large positive coefficients with large negative ones that
mostly cancel on the fitted values. The ridge penalty rules these out
and returns a smaller, more stable set of coefficients in exchange
for a small amount of bias. The parameter $\alpha$ sets the trade-off.
At $\alpha = 0$ Ridge reduces to OLS, and as $\alpha \to \infty$ the
coefficients are driven to zero. We search $\alpha$ on a 50-point
log-spaced grid in $[10^{-2}, 10^3]$.

Principal component regression replaces the 11 near-collinear year
columns with their first $k$ principal components and fits OLS in
that reduced space. The PC directions beyond the leading few carry
almost no variance in the GDP block, so including them in OLS would
inflate the fit's sensitivity to them without supplying any real
signal. Cutting the number of components at $k$ eliminates those
directions outright.

### Ridge and PCR as shrinkage on principal directions

Both estimators can be written as filters on the principal directions
of the predictor matrix. Let the SVD of the centered $X$ be
$$X = U D V^T,$$
where $D$ holds the singular values in decreasing order. The OLS fit
projects $y$ onto the column space of $X$, weighting every PC
direction equally, while Ridge and PCR replace those unit weights
with component-wise filters,
$$\hat y_{\text{OLS}} = \sum_i u_i u_i^T y, \qquad
  \hat y_{\text{Ridge}} = \sum_i \frac{d_i^2}{d_i^2 + \alpha} u_i u_i^T y, \qquad
  \hat y_{\text{PCR}} = \sum_{i=1}^{k} u_i u_i^T y.$$
Ridge's factor is nearly one on high-variance directions and near
zero on low-variance ones, giving a smooth shrinkage. PCR is the hard
version, keeping the first $k$ directions untouched and dropping the
rest. In the GDP block the first three singular values dominate the
remaining eight by orders of magnitude, so both filters produce fits
built almost entirely from the first three PC directions. Ridge's
nested-CV RMSE and PCR's are two smoothings of the same low-rank
structure, explaining why they land so close.

```{r reg-setup}
y_reg <- lifeExp[, "2007"]
X_raw <- as.matrix(gdp[, 1:11]); colnames(X_raw) <- paste0("y", colnames(X_raw))
X_log <- log(X_raw)
alpha_grid <- 10^seq(-2, 3, length.out = 50)
k_grid     <- 1:11

ridge_fit <- function(X, y, alpha) {
  mx <- colMeans(X); sx <- apply(X, 2, sd); my <- mean(y)
  Xs <- sweep(sweep(X, 2, mx, "-"), 2, sx, "/"); ys <- y - my
  list(mx=mx, sx=sx, my=my,
       beta_s = solve(crossprod(Xs) + alpha * diag(ncol(Xs)), crossprod(Xs, ys)))
}
ridge_predict <- function(fit, Xnew) {
  Xs <- sweep(sweep(Xnew, 2, fit$mx, "-"), 2, fit$sx, "/")
  as.numeric(Xs %*% fit$beta_s) + fit$my
}
pcr_fit <- function(X, y, k) {
  mx <- colMeans(X); sx <- apply(X, 2, sd)
  Xs <- sweep(sweep(X, 2, mx, "-"), 2, sx, "/")
  p <- prcomp(Xs, scale. = FALSE); S <- p$x[, 1:k, drop = FALSE]
  list(mx=mx, sx=sx, rot=p$rotation[, 1:k, drop = FALSE],
       lm_fit=lm(y ~ ., data = data.frame(y=y, S)))
}
pcr_predict <- function(fit, Xnew) {
  Xs <- sweep(sweep(Xnew, 2, fit$mx, "-"), 2, fit$sx, "/")
  as.numeric(predict(fit$lm_fit, newdata = data.frame(Xs %*% fit$rot)))
}
loocv_ols <- function(X, y) {
  sapply(seq_along(y), function(i) {
    m <- lm(y ~ ., data = data.frame(y=y[-i], X[-i, ]))
    predict(m, newdata = data.frame(t(X[i, ])))
  })
}
rmse <- function(a, b) sqrt(mean((a - b)^2))
r2   <- function(a, b) 1 - sum((a - b)^2) / sum((a - mean(a))^2)
```

```{r reg-nested, cache=TRUE}
nested_ridge <- function(X, y, grid, n_inner = 10) {
  n <- length(y); preds <- numeric(n); alphas <- numeric(n)
  for (i in 1:n) {
    Xtr <- X[-i, ]; ytr <- y[-i]; set.seed(i)
    fold <- sample(rep(1:n_inner, length.out = nrow(Xtr)))
    cv <- sapply(grid, function(a) mean(sapply(1:n_inner, function(k) {
      fit <- ridge_fit(Xtr[fold != k, ], ytr[fold != k], a)
      rmse(ytr[fold == k], ridge_predict(fit, Xtr[fold == k, , drop = FALSE]))
    })))
    a_best <- grid[which.min(cv)]; alphas[i] <- a_best
    preds[i] <- ridge_predict(ridge_fit(Xtr, ytr, a_best), X[i, , drop = FALSE])
  }
  list(preds = preds, alphas = alphas)
}
nested_pcr <- function(X, y, grid, n_inner = 10) {
  n <- length(y); preds <- numeric(n); ks <- integer(n)
  for (i in 1:n) {
    Xtr <- X[-i, ]; ytr <- y[-i]; set.seed(i)
    fold <- sample(rep(1:n_inner, length.out = nrow(Xtr)))
    cv <- sapply(grid, function(k) mean(sapply(1:n_inner, function(f) {
      fit <- pcr_fit(Xtr[fold != f, ], ytr[fold != f], k)
      rmse(ytr[fold == f], pcr_predict(fit, Xtr[fold == f, , drop = FALSE]))
    })))
    k_best <- grid[which.min(cv)]; ks[i] <- k_best
    preds[i] <- pcr_predict(pcr_fit(Xtr, ytr, k_best), X[i, , drop = FALSE])
  }
  list(preds = preds, ks = ks)
}
p_ols_raw <- loocv_ols(X_raw, y_reg)
p_ols_log <- loocv_ols(X_log, y_reg)
nr_raw <- nested_ridge(X_raw, y_reg, alpha_grid)
nr_log <- nested_ridge(X_log, y_reg, alpha_grid)
np_raw <- nested_pcr(X_raw,  y_reg, k_grid)
np_log <- nested_pcr(X_log,  y_reg, k_grid)
```

On log(GDP) the selected $\alpha$ is stable across the 141 outer
folds (range `r sprintf("%.1f–%.1f", min(nr_log$alphas), max(nr_log$alphas))`,
median `r sprintf("%.1f", median(nr_log$alphas))`). On raw GDP it is
less stable (range
`r sprintf("%.2f–%.1f", min(nr_raw$alphas), max(nr_raw$alphas))`,
median `r sprintf("%.1f", median(nr_raw$alphas))`), reflecting a flat
cross-validation surface on the raw scale where different subsets of
140 countries favour very different penalties.

```{r reg-path, fig.cap="Standardised Ridge coefficients as a function of the penalty $\\alpha$, on raw GDP (left) and log GDP (right). The vertical dashed line marks the median $\\alpha$ selected across the 141 nested-LOOCV outer folds.", fig.width=11, fig.height=4}
alphas_path <- 10^seq(-2, 3, length.out = 100)
coef_path <- function(X, y) {
  mx <- colMeans(X); sx <- apply(X, 2, sd)
  Xs <- sweep(sweep(X, 2, mx, "-"), 2, sx, "/"); ys <- y - mean(y)
  XtX <- crossprod(Xs)
  t(sapply(alphas_path, function(a)
    solve(XtX + a * diag(ncol(Xs)), crossprod(Xs, ys))))
}
cp_raw <- coef_path(X_raw, y_reg); cp_log <- coef_path(X_log, y_reg)
years_pred <- seq(1952, 2002, 5)
path_df <- rbind(
  data.frame(alpha = rep(alphas_path, 11),
             coef  = as.vector(cp_raw),
             year  = factor(rep(years_pred, each = length(alphas_path))),
             which = "raw GDP"),
  data.frame(alpha = rep(alphas_path, 11),
             coef  = as.vector(cp_log),
             year  = factor(rep(years_pred, each = length(alphas_path))),
             which = "log GDP"))
marks <- data.frame(which = c("raw GDP","log GDP"),
                    chosen = c(median(nr_raw$alphas), median(nr_log$alphas)))
ggplot(path_df, aes(alpha, coef, colour = year)) +
  geom_line(linewidth = 0.5) +
  geom_vline(data = marks, aes(xintercept = chosen),
             linetype = "dashed", colour = "red", linewidth = 0.5) +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.3) +
  scale_x_log10() + scale_colour_viridis_d(name = "Year") +
  facet_wrap(~which) +
  labs(x = expression(alpha), y = "Standardised coefficient") +
  theme_paper
```

At low $\alpha$ the eleven year coefficients are large and opposing
in sign. With near-collinear predictors, OLS can pair up large
coefficients that mostly cancel when predicting the response.
Increasing $\alpha$ shrinks them toward zero and flattens the
year-to-year differences. On log(GDP) the nested-CV median sits in
the intermediate region, where shrinkage stabilises the coefficients
without yet removing their signal.

```{r reg-table}
k_log <- as.integer(names(sort(table(np_log$ks), decreasing = TRUE))[1])
k_raw <- as.integer(names(sort(table(np_raw$ks), decreasing = TRUE))[1])
results <- data.frame(
  Model = c("OLS (raw GDP)", "OLS (log GDP)",
            "Ridge (raw GDP)", "Ridge (log GDP)",
            sprintf("PCR (raw GDP, k=%d)", k_raw),
            sprintf("PCR (log GDP, k=%d)", k_log)),
  `LOOCV RMSE` = sprintf("%.2f",
    c(rmse(y_reg, p_ols_raw), rmse(y_reg, p_ols_log),
      rmse(y_reg, nr_raw$preds), rmse(y_reg, nr_log$preds),
      rmse(y_reg, np_raw$preds), rmse(y_reg, np_log$preds))),
  `LOOCV R2` = sprintf("%.2f",
    c(r2(y_reg, p_ols_raw), r2(y_reg, p_ols_log),
      r2(y_reg, nr_raw$preds), r2(y_reg, nr_log$preds),
      r2(y_reg, np_raw$preds), r2(y_reg, np_log$preds))),
  check.names = FALSE)
kable(results,
      caption = "Nested leave-one-out cross-validated performance.",
      booktabs = TRUE)
```

OLS on raw GDP gives a LOOCV RMSE of
`r sprintf("%.2f", rmse(y_reg, p_ols_raw))` years and $R^2 =$
`r sprintf("%.2f", r2(y_reg, p_ols_raw))`, while OLS on log(GDP)
gives RMSE `r sprintf("%.2f", rmse(y_reg, p_ols_log))` and
$R^2 =$ `r sprintf("%.2f", r2(y_reg, p_ols_log))`, a
one-and-a-half-year reduction in prediction error from the transform.
Ridge on log(GDP) improves this further to RMSE
`r sprintf("%.2f", rmse(y_reg, nr_log$preds))` and $R^2 =$
`r sprintf("%.2f", r2(y_reg, nr_log$preds))`, while Ridge on raw GDP
reaches RMSE `r sprintf("%.2f", rmse(y_reg, nr_raw$preds))` and
$R^2 =$ `r sprintf("%.2f", r2(y_reg, nr_raw$preds))`, a modest gain
over the raw-scale OLS. GDP's skewness makes the linear relationship
with life expectancy non-linear on the raw scale but close to linear
on the log scale, and the regression results express the cost of
ignoring this as 0.19 units of $R^2$.

PCR has the lowest RMSE. On log(GDP) the inner CV picked $k = 3$
components in every one of the 141 outer folds, giving nested RMSE
`r sprintf("%.2f", rmse(y_reg, np_log$preds))` and $R^2 =$
`r sprintf("%.2f", r2(y_reg, np_log$preds))`. The first three GDP
components already explain 99% of the variance, so dropping any
single country never changes which directions dominate the design.
PCR on raw GDP peaks at $k = 2$ with RMSE
`r sprintf("%.2f", rmse(y_reg, np_raw$preds))`, still short of the
log-based models. The transform matters more than the
multicollinearity fix.

For PCR the held-out country can't influence the $k$ used to score
it, so the nested RMSE is an unbiased estimate of out-of-sample RMSE.
For Ridge the inner-selected $\alpha$ varies across outer folds, so
the nested score exceeds a flat LOOCV. Both outperform OLS on
log(GDP) by addressing the near-collinear year columns.

```{r reg-best, fig.cap="Leave-one-out predictions against actual 2007 life expectancy for the best model.", fig.width=7, fig.height=6}
best_idx <- which.min(c(rmse(y_reg, p_ols_raw), rmse(y_reg, p_ols_log),
                        rmse(y_reg, nr_raw$preds), rmse(y_reg, nr_log$preds),
                        rmse(y_reg, np_raw$preds), rmse(y_reg, np_log$preds)))
best_preds <- list(p_ols_raw, p_ols_log, nr_raw$preds, nr_log$preds,
                   np_raw$preds, np_log$preds)[[best_idx]]
pd <- data.frame(actual = y_reg, predicted = best_preds,
                 country = country, continent = continent)
pd$ra <- abs(pd$predicted - pd$actual)
worst3 <- pd[order(-pd$ra), ][1:3, ]
ggplot(pd, aes(actual, predicted, colour = continent)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 2, alpha = 0.75) +
  geom_text_repel(data = worst3, aes(label = country),
                  size = 3, fontface = "bold", show.legend = FALSE) +
  scale_colour_manual(values = palette_continents) +
  labs(x = "Actual 2007 life expectancy (years)",
       y = "LOOCV predicted life expectancy (years)",
       title = results$Model[best_idx], colour = "Continent") +
  theme_paper
```

```{r reg-residuals, include=FALSE}
resid_all <- best_preds - y_reg
sd_r <- sd(resid_all); thr <- 2 * sd_r
outliers <- which(abs(resid_all) > thr)
```

The 141 leave-one-out residuals have SD
`r sprintf("%.2f", sd_r)`. All `r length(outliers)` countries beyond
2 SD are African and over-predicted. Swaziland, Botswana, and South
Africa miss by more than 20 years, then Angola, Lesotho, Equatorial
Guinea, and Gabon by between 15 and 17. Zambia, Namibia, and Nigeria
sit just inside the threshold, extending the pattern.

These countries hit GDP-decoupled shocks the model can't see. The
southern African group lost life expectancy from the 1990s onward
without a proportional fall in income, an HIV-driven health collapse,
while the oil-rent economies (Equatorial Guinea, Gabon, Angola)
earned resource income concentrated in a narrow elite without
matching public-health gains. The two negative-residual outliers run
the opposite way, with Vietnam ($-14.2$) and Nicaragua ($-13.0$)
gaining life expectancy faster than GDP. Every combination of transform,
regularisation, and component count we tried left the same countries
furthest from the diagonal.
